### 2020-05-19 manuscript.tex
small edits

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. In ferns, deciphering species boundaries can be even more difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a very high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Otto2000, Otto2000}. Such genomic changes can contribute to cryptic variation by altering niche requirements or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet be essentially phenotypically invisible \autocite{Patel2019}.

@@

As a well-studied, pan-tropical group known to include cryptic species, \textit{Ceratopteris}, and in particular \textit{C. thalictroides}, is an ideal system in which to study the process of cryptic speciation. In this study, we produce the first multilocus genomic analysis of \textit{Ceratopteris} using restriction-site associated DNA sequencing (RADSeq). RADSeq is a cost-effective way to generate a large amount of genomic data, and many downstream analysis tools are currently available. We apply three approaches to identify cryptic species. First, we estimate population structure and hybridization; second, we reconstruct phylogenetic relationships among samples; finally, we investigate ploidy levels across individuals. This study is a step towards resolving species boundaries within this convoluted genus. Our work will also provide a phylogenetic reference for future studies on \textit{Ceratopteris}, as well as find areas within the genus in need of additional research.

@@

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADSeq to generate a large genomic dataset, which was processed using the ipyrad pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADSeq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

All parameter values and code for data processing and downstream analyses can be found on GitHub (github.com/sylviakinosian/ceratopteris\_RADSeq).

@@

We started with 90 samples, 56 from herbarium specimens and 34 from silica-dried tissue. These samples represented five of seven named species of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} from Australia for use as an outgroup. We also included a sample of the common lab strain of \textit{C. richardii} (\textit{Hnn}). The two missing species of \textit{Ceratopteris} were cryptic species of \textit{C. thalictroides} (L.) Brongn. \autocite{Masuyama2010}: \textit{C. oblongiloba} Masuyama \& Watano, known from Southeast Asia; and \textit{C. froesii} Brade, endemic to Brazil. Both have relatively narrow ranges and few herbarium species, especially \textit{C. froesii}. We were unable to obtain any material from herbarium specimens labeled as either of these species, and none of the plants collected in the field keyed out to \textit{C. oblongiloba} or \textit{C. froesii}.

@@

Raw data were demultiplexed using stacks v. 2.4 \autocite{Catchen2011, Catchen2013} process\_radtags, allowing for a maximum of one mismatch per barcode. Demultiplexed FASTQ files were paired merged using ipyrad version 0.7.30 \autocite{Eaton2020}. Low quality bases, plus adapters and primers were removed from each read. Filtered reads were clustered at the default setting of 85\%, and we required a sequencing depth of 6 or greater per base and a minimum of 30 samples per locus to be included in the final assembly. The ipyrad pipeline defines a locus as a short sequence present across samples. From each locus, ipyrad identifies single nucleotide polymorphisms (SNPs); these SNPs are the variation used in downstream analyses. Since stacks and ipyrad assume diploidy, we processed all samples as such. It is well-known that some species of \textit{Ceratopteris} are non-diploid \autocite{Adjie2007, Masuyama2010}, and so we explore the ploidy of each sample more thoroughly later in our analyses.

@@

%We included five specimens that were lower quality (see Supplementary data 1) because they were important in achieving a more even sampling distribution. They did not cluster with each other in our analyses, and their assignments in the population structure and phylogenetic analyses were biologically plausible.

@@

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium specimens, we could not perform flow cytometry or chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to detect ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADSeq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from ipyrad.

@@

\textbf{2.8. \textit{In silico} analysis of the \textit{Ceratopteris richardii} genome}

Our sampling included relatively few individuals of \textit{Ceratopteris richardii}, so we utilized the published genome from \textcite{Marchant2019} to perform an \textit{in silico} RADseq digest, creating a pseudo-indiviual to include in our analyses. We used the program ddRADSeqTools \autocite{Mora-Marquez2017} to digest the genome with the same enzymes used in our \textit{in vivo} digest (PstI and BfaI). Then we performed a simulated ddRADSeq run with paired-end reads, demultiplexing, and trimming using ddRADSeqTools. This simulated indidivuals was added to our other samples and processed via the ipyrad pipeline. Finally, we ran STRUCTURE for this set of individuals.

@@

\textbf{3.4. Analysis of simulated RADSeq data}

We attempted to create a pseudo-individual to including in our analysis using the \textit{Ceratopteris richardii} genome assembly. We were able to obtain 9,047,554 simulated raw sequences utilizing an \textit{in silico} RADSeq digest. These reads, in combination with the rest of our \textit{in vivo} data, were processed with the ipyrad pipeline. The ipyrad pipeline consists of 7 steps, each performing one aspect of RADSeq data processsing \autocite{Eaton2020}. Our pseduo-individual of \textit{C. richardii} performed nearly identically to all of the other samples until Step 6. Here, similar sequences are clustered across samples, and then these shared sequences are mapped to a reference (our reference was a \textit{de novo} assembly, because the quality and coverage of the \textit{C. richardii} genome was too poor to serve as a reference). This requires that all individuals are related relatively close to one another, so that they share a large number of sequences. Very few sequences were retained for our pseudo-individual of \textit{C. richardii}, indicating that the sequences obtained are too different from the \textit{in vivo} samples. In order to retain enough SNPs to run STRUCTURE, we had to reduce the minimum number of individuals to have data at a particular locus to just 3. In our entirely \textit{in vivo} dataset, this minimum number of individuals was 30; we designated such a high number to reduce missing data. In our dataset with the pseud-individual of \textit{C. richardii}, a vast majority of our retained loci had missing data.

The differences between the pseudo-individual derived from the \textit{Ceratopteris richardii} and our other samples is mostly likely due to the fact that the current assembly (v. 1.1) only represents about one third of the total genome, and consists of mostly short reads. Long reads that can span the many repetitive elements in the genome comprise only 0.03\% of the final assembly \autocite{Marchant2019}. Because of these factors, the current genome assembly contains small amount of short sequences that are difficult to accurately compare to the fragments generated by \textit{in vivo} RADSeq. We have included the results from our \textit{in silico} RADSeq digest in Appendix A: Supplementary data 2, but do not present them in the main manuscript.

>>>>

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. In ferns, deciphering species boundaries can be even more difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Sigel2016, Otto2000, Otto2000}. Changes in ploidy can contribute to cryptic variation by altering niche space or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet have small effects on phenotype \autocite{Patel2019}.

@@

As a well-studied, pan-tropical group known to include cryptic species, \textit{Ceratopteris}, and in particular \textit{C. thalictroides}, is an ideal system in which to study the process of cryptic speciation. In this study, we produce the first multilocus genomic analysis of \textit{Ceratopteris} using restriction-site associated DNA sequencing (RADseq). RADseq is a cost-effective way to generate a large amount of genomic data, and many downstream analysis tools are currently available. We apply three approaches to identify cryptic species. First, we estimate population structure and hybridization; second, we reconstruct phylogenetic relationships among samples; finally, we investigate ploidy levels across individuals. This study is a step towards resolving species boundaries within this convoluted genus. Our work will also provide a phylogenetic reference for future studies on \textit{Ceratopteris}, as well as find areas within the genus in need of additional research.

@@

To best analyze the cryptic diversity within \textit{Ceratopteris} we gathered samples from across its pan-tropical distribution. We utilized RADseq to generate a large genomic dataset, which was processed using the ipyrad pipeline \autocite{Eaton2020} for downstream analysis. Our three-part approach for assessing cryptic diversity takes advantage of the flexibility of RADseq data, and provides a unique window into the processes of speciation and diversification in a complex genus.

All parameter values and code for data processing and downstream analyses can be found on GitHub (github.com/sylviakinosian/ceratopteris\_RADseq).

@@

We collected 90 samples, 56 from herbarium specimens and 34 from silica-dried tissue. These samples represented five of seven named species of \textit{Ceratopteris}, and one sample of \textit{Acrostichum aureum} from Australia for use as an outgroup. Included in our silica-dried specimens was a sample of the common lab strain of \textit{C. richardii} (\textit{Hnn}). The two missing species of \textit{Ceratopteris} were cryptic species of \textit{C. thalictroides} (L.) Brongn. \autocite{Masuyama2010}: \textit{C. oblongiloba} Masuyama \& Watano, known from Southeast Asia; and \textit{C. froesii} Brade, endemic to Brazil. Both have relatively narrow ranges and few herbarium species, especially \textit{C. froesii}. We were unable to obtain any material from herbarium specimens labeled as either of these species, and none of the plants collected in the field keyed out to \textit{C. oblongiloba} or \textit{C. froesii}.

@@

Raw data were demultiplexed using stacks v. 2.4 \autocite{Catchen2011, Catchen2013} process\_radtags, allowing for a maximum of one mismatch per barcode. Demultiplexed FASTQ files were paired merged using ipyrad version 0.9.52 \autocite{Eaton2020}. Low quality bases, plus adapters and primers were removed from each read. Filtered reads were clustered at the default setting of 85\%, and we required a sequencing depth of 6 or greater per base and a minimum of 30 samples per locus to be included in the final assembly. The ipyrad pipeline defines a locus as a short sequence present across samples. From each locus, ipyrad identifies single nucleotide polymorphisms (SNPs); these SNPs are the variation used in downstream analyses. Since stacks and ipyrad assume diploidy, we processed all samples as such. It is well-known that some species of \textit{Ceratopteris} are non-diploid \autocite{Adjie2007, Masuyama2010}, and so we explore the ploidy of each sample more thoroughly later in our analyses.

@@

@@

% NOTE: add in mrbayes, raxml, and/or bucky here

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium specimens, we could not perform flow cytometry or chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to detect ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADseq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from ipyrad.

@@

\textbf{2.8. Simulated RADseq analysis of the \textit{Ceratopteris richardii} genome}

Our sampling included relatively few individuals of \textit{Ceratopteris richardii}, so we utilized the published genome from \textcite{Marchant2019} to perform an \textit{in silico} RADseq digest, creating a pseudo-indiviual to include in our analyses. We used the program ddRADseqTools \autocite{Mora-Marquez2017} to digest the genome with the same enzymes used in our \textit{in vivo} digest (PstI and BfaI). Then we performed a simulated ddRADseq run with paired-end reads, demultiplexing, and trimming using ddRADseqTools. This simulated individual was added to our other samples, processed via the ipyrad pipeline, and analyzed via STRUCTURE.

@@

\textbf{3.4. Analysis of simulated RADseq data}

We attempted to create a pseudo-individual to include in our analysis using the \textit{Ceratopteris richardii} genome assembly. We were able to obtain 9,047,554 simulated raw sequences utilizing an \textit{in silico} RADseq digest. These reads, in combination with the rest of our \textit{in vivo} data, were processed with the ipyrad pipeline. The ipyrad pipeline consists of 7 steps, each performing one aspect of RADseq data processsing \autocite{Eaton2020}. Our pseduo-individual of \textit{C. richardii} performed nearly identically to all of the other samples until Step 6. Here, similar sequences are clustered across samples, and then these shared sequences are mapped to a reference (our reference was a \textit{de novo} assembly, because the quality and coverage of the \textit{C. richardii} genome was too poor to serve as a reference). This requires that all individuals are related relatively close to one another, so that they share a large number of sequences. Very few sequences were retained for our pseudo-individual of \textit{C. richardii}, indicating that the sequences obtained are too different from the \textit{in vivo} samples. In order to retain enough SNPs to run STRUCTURE, we required that only 3 individuals needed to have data at a given locus for it to be included in the final data set. In our entirely \textit{in vivo} dataset, this minimum number of individuals was 30; we designated such a high number to reduce missing data. In our dataset with the pseud-individual of \textit{C. richardii}, a vast majority of our retained loci had missing data.

The differences between the pseudo-individual derived from the \textit{Ceratopteris richardii} and our other samples is mostly likely due to the fact that the current assembly (v. 1.1) only represents about one third of the total genome, and consists of mostly short reads. Long reads that can span the many repetitive elements in the genome comprise only 0.03\% of the final assembly \autocite{Marchant2019}. Because of these factors, the current genome assembly contains a small amount of short sequences that are difficult to accurately compare to the fragments generated by \textit{in vivo} RADseq. We have included the results from our \textit{in silico} RADseq digest in Appendix A: Supplementary data 2, but do not present them in the main manuscript.
